---
title: 'Assignment 6: Web APIs'
author: "Prinon Mahdi"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(jsonlite)
library(tidyverse)
```

I have decided to read in data from the **Books API**. In particular, I will pull the most recent bestsellers list for fiction (both print and ebook).

```{r}
link <- 'https://api.nytimes.com/svc/books/v3/lists/current/combined-print-and-e-book-fiction.json?api-key=EtTLK80hGeDW3v0ipV1Rmr5tBL7V1rSk'
json_list <- fromJSON(link) 
current_fictions <- as_tibble(json_list$results$books)
glimpse(current_fictions)
```

It is no use cleaning the data because this bestsellers list is updated weekly. A column devoid of any information this week might have relevant information next week, so it can't be dropped. However, it makes sense to create a subset of the main dataframe so that only the essential information (title, author, and description) is readily available for quick perusal.

```{r}
basic_info <- current_fictions %>% 
  select(title, author, description)
knitr::kable(basic_info)
```

Overall, this notebook can be run weekly to get the up-to-date bestsellers list for fiction from The New York Times.