---
title: "Assignment 2 | SQL and R"
author: "Prinon Mahdi"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

The database I will be connecting to is hosted on Google Cloud. To connect to it, several packages need to be loaded.

```{r library}
library('tidyverse')
library('DBI')
library('RPostgres')
```

Now, let's set up the connection.

```{r connecting-to-google-cloud}
con <- dbConnect(
  Postgres(),
  dbname = 'postgres',
  host = '34.132.103.49',
  port = '5432',
  user = 'postgres',
  password = 'postgres')
```

Checking to see if it worked:

```{r test-query}
  dbGetQuery(
    con, 
    'SELECT *
    FROM movies;')
```

It worked! Now, I want to create a single data frame by joining data from both tables in the database.

```{r inner-join-movies-viewers}
joined_df <- 
  dbGetQuery(
    con,
    'SELECT * FROM viewers
    INNER JOIN movies
    ON viewers.movie_id = movies.id;'
  )
glimpse(joined_df)
```
It seems there are **NA** values in the *rating* column of the data frame. I will resolve this issue in 2 different ways. Firstly, I will remove the **NA** values in R. Secondly, I will modify my query so that the joined data excludes records with **NA** values. In my opinion, excluding the records would be a valid approach since the missing information indicates that the specific viewer did not watch a particular movie, so he or she was unable to rate it.

```{r removing-na-in-r}
joined_df <- na.omit(joined_df)

# Checking that no NA values remain
all(complete.cases(joined_df)) == TRUE
```

Next, I will write a new query to retrieve the data from the database while excluding **NA** records and adding a new column called *adjusted_rating* that scales *rating* values by 2 so that *rating* and *imdb_rating* are comparable.

```{r excluding-na-adjusted_rating}
df_no_na <-     
  dbGetQuery(
    con,
    'SELECT 
    index, 
    name, 
    movie_id, 
    rating, 
    id, 
    title, 
    release_yr, 
    imdb_rating,
    rating * 2 AS adjusted_rating
    FROM viewers
    INNER JOIN movies
    ON viewers.movie_id = movies.id
    WHERE rating IS NOT NULL;')

# Checking that no NA values remain
all(complete.cases(df_no_na)) == TRUE

# Checking that adjusted_rating column has been added
glimpse(df_no_na)
```
