---
title: 'Assignment 5: Working with XML and JSON in R'
author: "Prinon Mahdi"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(xml2)
library(rjson)
library(htmltab)
```

# XML

First step is reading in the xml data, and then extracting the relevant information from various nodes by navigating with xpath.

```{r read-parse-xml}
xml_file_url <- 'https://raw.githubusercontent.com/pmahdi/cuny-data-607/main/assignment-5-books.xml'
xml_tree <- read_xml(x = xml_file_url)

all_titles_atomic <- xml_text(
  xml_find_all(
    xml_tree, '/books/book/title'
    )
  )

all_authors_atomic <- xml_text(
  xml_find_all(
    xml_tree, '/books/book/author'
    )
  )

book1_chapters_atomic <- xml_text(
  xml_find_all(
    xml_tree, '/books/book[1]/chapters/chapter'
    )
  )

book2_chapters_atomic <- xml_text(
  xml_find_all(
    xml_tree, '/books/book[2]/chapters/chapter'
    )
  )

book3_chapters_atomic <- xml_text(
  xml_find_all(
    xml_tree, '/books/book[3]/chapters/chapter'
    )
  )

all_pages_atomic <- xml_text(
  xml_find_all(
    xml_tree, '/books/book/pages'
    )
  )

all_formats_atomic <- xml_text(
  xml_find_all(
    xml_tree, '/books/book/format'
    )
  )
```

With the messy part completed, now it's a matter of combining and organizing the information in a tidy way. To that end, first I am going to create a tibble that gives basic information on each of the 3 titles.

```{r xml-tibble-books}
xml_books <- tibble(
  title = all_titles_atomic, 
  authors = c(2, 1, 1), 
  chapters = c(length(book1_chapters_atomic), 
               length(book2_chapters_atomic), 
               length(book3_chapters_atomic)
               ), 
  pages = as.numeric(all_pages_atomic), 
  format = all_formats_atomic
  )
```

Next, I will create a tibble that pairs the authors with their respective titles.

```{r xml-tibble-authors}
xml_authors <- tibble(
  author = c(
    all_authors_atomic[1], 
    all_authors_atomic[4], 
    all_authors_atomic[2], 
    all_authors_atomic[3]
    ), 
  title = c(
    all_titles_atomic[1], 
    all_titles_atomic[3], 
    all_titles_atomic[1], 
    all_titles_atomic[2]
    )
  )
```

Finally, all that's left is to create a tibble for pairing titles with their corresponding chapters.

```{r xml-tibble-chapters}
xml_chapters_book1 <- tibble(
  title = all_titles_atomic[1], 
  chapter = book1_chapters_atomic
)

xml_chapters_book2 <- tibble(
  title = all_titles_atomic[2], 
  chapter = book2_chapters_atomic
)

xml_chapters_book3 <- tibble(
  title = all_titles_atomic[3], 
  chapter = book3_chapters_atomic
)

xml_chapters <- bind_rows(xml_chapters_book1, 
                          xml_chapters_book2, 
                          xml_chapters_book3)
```

The three tibbles `xml_books`, `xml_authors`, and `xml_chapters` present the data stored in the original xml file in a tidy way. I have tried to stick to the philosophy of having each column as a variable, each row as an observation, and each cell as a single value. That is why I have felt it necessary to split the data into 3 separate tibbles. Let's have a peek at the final result:

```{r xml-final}
xml_books
xml_authors
glimpse(xml_chapters)
# Showing the first and last 5 rows of xml_chapters
xml_chapters %>% 
  slice(c(head(row_number(), 5), tail(row_number(), 5)))
```

# JSON

First and foremost, the json file has to be read into the R environment.

```{r read-json}
json_file_url <- 'https://raw.githubusercontent.com/pmahdi/cuny-data-607/main/assignment-5-books.json'
json_list <- fromJSON(file = json_file_url)
```

As seen previously, the information I have chosen to include about my books requires me to create multiple tibbles to present the data in a tidy way. However, it **is** possible to organize the information in a single data structure if the tidy philosophy is disregarded. For example: 

```{r json-dirty}
json_dirty <- sapply(json_list, `[`)
knitr::kable(json_dirty)
```

This is an extremely ugly way to organize and display the information. So, I will now tidy the data step by step as I did in the XML section.

```{r json-dirty-parse}
# Getting all the titles in one character vector
json_title_atomic <- character()
for (i in c(1, 6, 11)) {
  json_title_atomic <- append(x = json_title_atomic, values = json_dirty[[i]])
}
# Getting all the authors in one character vector
json_author_atomic <- character()
for (i in c(2, 7, 12)) {
  json_author_atomic <- 
    append(x = json_author_atomic, values = json_dirty[[i]])
}
json_author_atomic[c(2, 4)] <- json_author_atomic[c(4, 2)]
# Extracting the chapters of each book separately, then matching them with their corresponding title
json_book_1_chapters <- tibble(title = json_title_atomic[1], 
                               chapter = json_dirty[[3]])
json_book_2_chapters <- tibble(title = json_title_atomic[2], 
                               chapter = json_dirty[[8]])
json_book_3_chapters <- tibble(title = json_title_atomic[3], 
                               chapter = json_dirty[[13]])

# Getting all the page counts in one character vector
json_pages_atomic <- numeric()
for (i in c(4, 9, 14)) {
  json_pages_atomic <- 
    append(x = json_pages_atomic, values = json_dirty[[i]])
}
# Getting all the publication formats in one character vector
json_format_atomic <- character()
for (i in c(5, 10, 15)) {
  json_format_atomic <- 
    append(x = json_format_atomic, values = json_dirty[[i]])
}
```

With all the necessary pieces parsed, it's time to recombine them to create nice and tidy tibbles.

```{r json-tidy-tibbles}
json_books <- tibble(
  title = json_title_atomic, 
  authors = c(2, 1, 1), 
  chapters = c(length(json_dirty[[3]]), length(json_dirty[[8]]), 
               length(json_dirty[[13]])), 
  pages = json_pages_atomic, 
  format = json_format_atomic
)

json_authors <- tibble(
  author = json_author_atomic, 
  title = c(json_title_atomic[1], json_title_atomic[3], json_title_atomic[2], 
            json_title_atomic[1])
)

json_chapters <- bind_rows(json_book_1_chapters, 
                          json_book_2_chapters, 
                          json_book_3_chapters)

json_books
json_authors
glimpse(json_chapters)
# Showing the first and last 5 rows of json_chapters
json_chapters %>%
  slice(c(head(row_number(), 5), tail(row_number(), 5)))
```

# HTML

Reading HTML data into R and creating tidy data structures is far easier than doing so with either JSON or XML. A very convenient package `htmltab` exists for this purpose, and I plan to use it, even though doing so almost feels like cheating because it's so easy.

```{r html-reading-in-dataframe}
html_file_url = 'https://raw.githubusercontent.com/pmahdi/cuny-data-607/main/assignment-5-books.html'
html_books <- as_tibble(htmltab(doc = html_file_url, which = 1))
html_authors <- as_tibble(htmltab(doc = html_file_url, which = 2))
html_chapters <- as_tibble(htmltab(doc = html_file_url, which = 3))

# Adding some extra information to html_books and shuffling the column order of html_authors
html_books <- html_books %>% 
  mutate(authors = c(2, 1, 1), chapters = c(11, 23, 23), .after = title)
html_authors[c(1, 2)] <- html_authors[c(2, 1)]
colnames(html_authors) <- c('author', 'title')

html_books
html_authors
glimpse(html_chapters)
# Showing the first and last 5 rows of html_chapters
html_chapters %>%
  slice(c(head(row_number(), 5), tail(row_number(), 5)))
```