---
title: "DATA 607 | Project 1"
author: "Prinon Mahdi"
output: html_document
---

```{r setup, include=FALSE, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Loading packages and reading in the data

```{r loading-reading}
library('tidyverse')

link <- 'https://raw.githubusercontent.com/pmahdi/cuny-data-607/main/proj-1-chess.txt'
chess_og <- read_delim(
  file = link, 
  delim = '|', 
  col_names = FALSE, 
  na = c('-----------------------------------------------------------------------------------------', ''), 
  trim_ws = TRUE,
  skip = 1, 
  skip_empty_rows = TRUE)

glimpse(chess_og)
```

### Transforming the data

It looks like column `X11` consists of only NA values. This is probably because the input file has a '|' at the end of each line, which also happens to be the delimiter character. So, *read_delim()* mistakenly adds on an extra column.

```{r removing-X11}
# Checking there are no values in X11
sum(!is.na(chess_og$X11))

# Dropping the column X11
chess_mod <- chess_og[-11]
glimpse(chess_mod)
```

It also seems like every 3rd row is full of NA values. So, let's drop all rows that only have NA values.

```{r dropping-only-na-rows}
all_na_index <- logical(length = nrow(chess_mod)) 

for (i in 1:nrow(chess_mod)) {
  all_na_index[i] <- all(is.na(chess_mod[i, ]))
}

chess_mod <- chess_mod[!all_na_index, ]
```

Next, new column names are added and the first 2 rows are dropped.

```{r add-colnames-drop-rows-1-2}
colnames(chess_mod) <- c('id', 'name', 'points', 'round_1', 'round_2', 'round_3', 'round_4', 'round_5', 'round_6', 'round_7')

chess_mod <- chess_mod[c(-1, -2), ]
```

The next step should be to add 2 extra columns to the data frame for populating with useful information from each player's 2nd row (state name and pre-tournament rating). After information from the second row of each player is added to the 2 columns, each player's second row (so, all the even rows) will be dropped.

```{r cbind-empty-cols}
chess_mod <- cbind(chess_mod, tibble(state = rep('', nrow(chess_mod)), pre_rating = rep('', nrow(chess_mod))))

for (i in seq(2, nrow(chess_mod), 2)) {
  chess_mod[i - 1, 'state'] <- chess_mod[i, 'id']
  chess_mod[i - 1, 'pre_rating'] <- chess_mod[i, 'name']
}

chess_mod <- chess_mod[-seq(2, nrow(chess_mod), 2), ]

# turns out R automatically assigns rownames corresponding to the previous indices
head(rownames(chess_mod))
rownames(chess_mod) <- NULL
head(rownames(chess_mod))  # much better, less confusing row indices
```

At this stage, the `pre_rating` column needs to be cleaned so that it contains only the pre-tournament rating of each player. For this, regular expressions will be used alongside the **stringr** package from tidyverse. The plan is to extract the correct information from each row using pattern matching and save that to a vector. Then, if the result looks good, that vector, after coercion to numeric, will update the information in the `pre_rating` column.

```{r clean-pre_rating}
extracted <- str_extract(string = chess_mod$pre_rating, pattern = ' \\d{3,4}')

# Checking that the extraction went alright
length(extracted) == nrow(chess_mod); sum(is.na(extracted))
head(extracted); tail(extracted)

# Need to remove the whitespace in front of each rating
extracted <- str_trim(string = extracted, side = 'both')

# Converting to numeric
extracted <- as.numeric(extracted)

# Populating chess_mod$pre_rating with extracted
chess_mod$pre_rating <- extracted
```

Almost all of the required information has been isolated in individual columns. Only average pre-tournament opponent rating is left. To get that value, the ids of each player's opponents have to be isolated.

```{r isolate-opponent-ids}
opp_id <- apply(X = chess_mod[4:10], MARGIN = 2, FUN = function(x) {str_extract(string = x, pattern = '\\d{1,2}')})

colnames(opp_id) <- c('opp_id_r1', 'opp_id_r2', 'opp_id_r3', 'opp_id_r4', 'opp_id_r5', 'opp_id_r6', 'opp_id_r7')

chess_mod <- cbind(chess_mod, opp_id)
```

After isolating the ids, a new numeric column named `avg_opp_pre_rating` is added to `chess_mod`. It's filled with 0s. The pre-tournament ratings of each player's opponents are fetched and those values are put through *mean()* to get the average opponent rating. Then, this output updates the `avg_opp_pre_rating` column for the corresponding player.

```{r calculating-avg_opp_pre_rating}
# Creating the column for average opponent rating and filling it with 0
chess_mod$avg_opp_pre_rating <- rep(0, nrow(chess_mod))

# Using a loop to calculate the avg_opp_pre_rating value for each player
for (i in 1:nrow(chess_mod)) {
  chess_mod[i, 'avg_opp_pre_rating'] <- 
    round(
      mean(
        chess_mod[
          as.character(chess_mod[i, 13:19]), 
          'pre_rating'], 
        na.rm = TRUE
        )
      )
}
```

Finally, all the required information has been isolated in individual columns. Those columns will be subsetted and saved to a new data frame named `chess_output`.

```{r creating-output-dataframe}
chess_output <- chess_mod[, c('name', 'state', 'points', 'pre_rating', 'avg_opp_pre_rating')]

colnames(chess_output) <- c('name', 'state', 'total_points', 'player_rating', 'avg_opp_rating')

# Converting total_points column to a numeric
chess_output$total_points <- as.numeric(chess_output$total_points)
glimpse(chess_output)
```
The final task is exporting `chess_output` to a csv file. The **readr** package's *write_csv* function can be used. 

```{r exporting-chess_output}
write_csv(
  x = chess_output, 
  file = './chess_output.csv', 
  na = 'NA', 
  append = FALSE, 
  col_names = TRUE, 
  quote = 'needed', 
  escape = 'double', 
  eol = '\n'
)
```